---
title: "sleep efficiency"
author: Sejal Davla, PhD
format: html
editor: visual
---

## Sleep Efficiency Dataset

### Data source: https://www.kaggle.com/datasets/equilibriumm/sleep-efficiency/code



```{r}
#Data import 
library(readr) 
rawdata <- read_csv("/Users/sejaldavla/Desktop/Data_Portfolio/Sleep/Sleep_Efficiency.csv")
rawdata
```
```{r}
#Required packages for data Cleaning, transformation and visualization 
library(dplyr) 
library(ggplot2) 
library(tidyr) 
library(RColorBrewer) 
library(forcats)
```


## Check the data structure, data cleanliness, and distribution
```{r}
str(rawdata)
sum(is.na(rawdata))
NA_data <- rawdata[!complete.cases(rawdata), ]
NA_data #Check for each column. No NA values in sleep amount, efficiency, and sleep cycle columns
```

## Checking data distribution
```{r}
A <- as.data.frame(rawdata %>%
  group_by(Age) %>%
  count(Gender))
A

his <- A %>%
  ggplot(aes(factor(Age), n, fill = Gender)) +
  geom_col(color = "#444443",alpha = 0.7, 
           position = position_dodge(preserve = "single")) +
  scale_y_continuous(limits = c(0,20)) +
  labs(
    title = "Sample distibution by age and gender in the sleep data",
    x = "Age",
    y = "number of people"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(size = 20),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 15)
  )
his
```
```{r fig.width= 6, fig.height=5}
plot <- rawdata %>%
  ggplot(aes(x = Gender, y = `Sleep duration`)) +
  geom_violin(width = 0.7, trim = FALSE, fill = "#E6E5E2") +
  geom_point(aes(color = factor(Age < 18)), 
             position = position_jitter(), size = 2, shape = 19, alpha = 0.7) +
  geom_boxplot(color = "black", width = 0.1, alpha = 0.2, outlier.colour = "#F8B31F") +
  scale_color_manual(values = c("#F8B31F", "#8107F4"),
                     labels = c("18 and Above", "Below 18")) +
  scale_y_continuous(limits = c(4, 11),
                     breaks = seq(4,11, by = 1)) +
  labs(x = NULL,
       y = "Sleep duration at night (Hours)",
       color = "Age") +
  theme_classic() +
  theme(
    axis.line.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_text(family = "Arial", size = 12),
    axis.text.x = element_text(size = 15, color = "blue")
  ) 
plot
```

## According to human sleep literature, sleep patterns vary between teenagers and adults. There are only 9 individuals below the age of 18 in this dataset. These data points are removed from further analysis to focus on the adult sleep patterns.

```{r}
summary <- rawdata %>%
  filter(Age >= 18) %>%
  group_by(Gender) %>%
  summarise(mean = mean(`Sleep duration`),
            sd = sd(`Sleep duration`),
            variance = var(`Sleep duration`))
summary
```
### Understanding sleep architecture in individuals. 
#Throughout the night, sleep cycles through different stages, namely, light sleep, deep sleep, and REM (rapid eye movement) sleep. The following graph shows individual differences in sleep stages. 

```{r}
#### Plotting data using sleep percentages
New <- rawdata %>%
  select(ID, `REM sleep percentage`,`Deep sleep percentage`,`Light sleep percentage`) %>%
  pivot_longer(!ID, names_to = "sleep percent", values_to = "percent") %>%
  rename(person = ID)

New %>%
  ggplot(aes(x = factor(person), y = percent, fill = `sleep percent`)) +
  geom_bar(position = "stack", stat = "identity") +
  scale_y_continuous(limits = c(0, 100),
                     breaks = seq(0, 100, by = 10)) +
  coord_flip(expand = FALSE, clip = "off") +
  theme(
    axis.text.y = element_blank(),
    axis.line.y = element_blank()
  )
```


## While sleep pecentages show proportion of individual sleep stages, it does not show them relative to the person's sleep amount. 


```{r}
#Convert sleep stage percentages to duration in hours
rawdata <- rawdata %>%
  mutate(REM = (`Sleep duration` * `REM sleep percentage`)/100) %>%
  mutate(Deep_sleep = (`Sleep duration` * `Deep sleep percentage`) / 100) %>%
  mutate(Light_sleep = (`Sleep duration` * `Light sleep percentage`) / 100)

```


```{r}
#### Plotting data using sleep amount
New_amount <- rawdata %>% 
  select(ID, Light_sleep, Deep_sleep, REM) %>%
  pivot_longer(!ID, names_to = "sleep amount", values_to = "values") %>%
  rename(person = ID)

baseplot <- New_amount %>%
  ggplot(aes(x = reorder(factor(person), -values), y = values, 
             fill = factor(`sleep amount`, 
                           levels = c("REM","Deep_sleep","Light_sleep")))) +
  geom_bar(position = "stack", stat = "identity") +
  scale_fill_manual(labels = c("REM", "Deep sleep","Light sleep"),
                    values = c("#D28ECA","#F4F6AF","#656ABB")) +
  scale_y_continuous(limits = c(0, 10),
                     breaks = seq(0, 10, by = 1)) +
  labs(
    y = "Sleep Amount (Hours)",
    x = NULL, 
    fill = "sleep cycle",
    title = "Sleep cycle variation in human population",
    subtitle = "While the total night sleep varies a lot from person to person, the REM sleep proportion \nremains unchanged.",
    caption = "By Sejal Davla, PhD"
    ) + 
  coord_flip(expand = FALSE, clip = "off") +
  theme_bw() +
  theme(
    plot.title = element_text(family = "Avenir", size = 20, color = "#472717"),
    plot.subtitle = element_text(family = "Avenir", size = 10, color = "#24788F"),
    axis.text.y = element_blank(),
    axis.line.y = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = c(0.89, 0.78),
    plot.background = element_rect(colour = "white"),
    panel.grid = element_blank(),
    axis.text.x = element_text(family = "Verdana", color = "#282130", size = 10),
    legend.text = element_text(family = "Verdana", color = "#282130", size = 8),
    legend.title = element_text(family = "Verdana", size = 10),
    legend.key.size = unit(0.5, "cm"),
    legend.box.background = element_rect(color = "grey", size = 2),
    ) +
  guides(fill = guide_legend(reverse = TRUE))

baseplot

ggsave("baseplot.png", units = "cm", height = 10, width = 15, dpi = 600)
```

